---
title: "Final-RMD-Project[Bird]"
author: "Roman Matthew Bird"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

# ABSTRACT
  Chronic Obstructive Pulmonary Disease (COPD) and air quality have been extensively studied to explore their interconnection. Our objective is to investigate this relationship by addressing the question: Does decreased air quality contribute to an increase in COPD cases across the United States? To achieve this, we analyzed air quality data from the World Population Review alongside COPD rates provided by the American Lung Association. Utilizing statistical methods, including the Pearson correlation test and generalized linear models, we sought to identify any associations between the two datasets. Our findings indicated a modest yet significant relationship, with 11% of COPD cases linked to prolonged exposure to poor air quality. Therefore, it is evident that exposure to low air quality can lead to COPD and other pulmonary diseases, underscoring the need for measures to improve air quality in order to enhance life expectancy in the United States.


# INTRODUCTION
 The Air Quality Index (AQI) is a standardized system designed to assess the quality of the air we breathe. It is determined by measuring four major air pollutants outlined by the Clean Air Act: ground-level ozone, particle pollution, carbon monoxide, and sulfur dioxide. The AQI is presented on a scale from 0 to 500, where 0 indicates the cleanest air and 500 signifies the most hazardous conditions. This scale effectively categorizes air quality, with lower values representing higher quality and higher values indicating poorer quality. The AQI has become a valuable tool that is widely integrated into weather applications, making it easily accessible to the general public.

Chronic obstructive pulmonary disease (COPD) is one of the leading causes of death in urban areas, and its development can be attributed to various factors such as tobacco use, occupational hazards, infections, and air pollution. This disease progresses gradually and tends to worsen with ongoing exposure to harmful substances. Unfortunately, the damage caused by COPD is typically irreversible and can ultimately lead to pulmonary failure. Several factors influence the prevalence of COPD, including geographical region, age, and gender.
  
STUDY QUESTION & HYPOTHESIS:

  With this information in mind, we have formulated the following question: “Do regions with poor air quality experience higher rates of Chronic Obstructive Pulmonary Disease (COPD)?” While numerous factors contribute to the development of COPD, a negative correlation between air quality and COPD rates could highlight several preventative measures that can be implemented to enhance lung health and mitigate the risk of the disease. These insights could influence health practices and empower individuals at risk for COPD to steer clear of environmental hazards, such as low air quality. Although various elements can affect COPD prevalence, we hypothesize that such a negative correlation does exist, and that areas characterized by lower air quality will indeed have higher rates of Chronic Obstructive Pulmonary Disease.



# METHODS
We collected data from two databases, World Population Review air quality database, and The National Lung Association COPD rates. All data was formatted, normalized, and tested in R. We ran the data through two statistical tests, a Pearson correlation test, and a generalized linear model.




```{r setup, include=FALSE}
#packages
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(viridis)
library(maps)
library(mapproj)
```

```{r data chunk, include = FALSE}

data_text <- "
StatePercentCountPercentCountPercentCount
Alabama 8.6 161,900 10.1 207,500 9.4 369,400
Alaska 5.3 15,300 6.1 16,100 5.7 31,400
Arizona 5.0 137,800 6.4 180,700 5.7 318,500
Arkansas 8.6 97,600 10.5 125,600 9.6 223,200
California 4.5 680,600 4.7 733,100 4.6 1,413,800
Colorado 4.3 98,900 5.9 135,100 5.1 233,900
Connecticut 4.7 65,400 5.9 87,200 5.3 152,500
Delaware 5.9 22,400 6.6 27,300 6.2 49,700
District of Columbia 4.1 10,600 5.1 15,100 4.6 25,700
Florida (2020) 6.8 574,600 8.2 745,500 7.5 1,320,100
Georgia 5.3 211,000 7.9 341,100 6.6 552,100
Hawaii 3.4 19,300 3.6 20,800 3.5 40,200
Idaho 5.0 35,400 6.4 45,800 5.7 81,200
Illinois 5.4 260,600 5.6 287,300 5.5 547,900
Indiana 7.4 188,300 9.5 256,200 8.5 444,400
Iowa 6.0 72,500 7.1 89,000 6.6 161,600
Kansas 5.9 64,400 7.0 79,300 6.4 143,700
Kentucky 9.8 166,800 12.0 214,900 10.9 381,700
Louisiana 8.5 144,500 8.9 163,700 8.7 308,200
Maine 8.1 43,600 9.8 56,200 9.0 99,900
Maryland 3.7 85,000 5.9 149,000 4.8 234,000
Massachusetts 5.0 136,300 6.7 196,200 5.9 332,500
Michigan 6.8 261,800 8.6 348,500 7.7 610,200
Minnesota 5.0 108,600 4.4 98,700 4.7 207,400
Mississippi 7.6 82,500 10.4 122,400 9.1 204,900
Missouri 6.6 153,500 10.4 255,600 8.5 409,000
Montana 5.2 22,800 6.6 28,400 5.9 51,200
Nebraska 5.2 38,300 6.1 45,600 5.6 83,900
Nevada 5.7 69,200 8.0 98,600 6.8 167,800
New Hampshire 4.7 165,200 6.3 235,000 5.5 400,100
New Jersey 6.3 35,100 8.1 46,400 7.2 81,400
New Mexico 4.5 36,100 6.7 55,900 5.6 91,900
New York 5.0 379,600 5.9 490,200 5.5 869,900
North Carolina 6.1 241,000 9.0 386,500 7.6 627,500
North Dakota 4.4 13,500 5.3 15,600 4.8 29,100
Ohio 7.6 340,300 9.5 452,900 8.6 793,200
Oklahoma 6.6 98,500 8.9 136,800 7.8 235,300
Oregon 5.1 85,200 7.0 120,300 6.1 205,500
Pennsylvania 6.9 344,000 6.8 361,300 6.8 705,300
Rhode Island 5.8 24,600 5.7 26,000 5.7 50,600
South Carolina 6.7 130,600 8.3 177,200 7.5 307,800
South Dakota 5.7 19,300 6.2 20,800 6.0 40,100
Tennessee 9.7 254,600 11.1 311,100 10.4 565,700
Texas 5.2 570,900 6.8 771,100 6.0 1,342,000
Utah 3.9 46,500 4.7 56,300 4.3 102,800
Vermont 5.8 15,000 7.3 19,700 6.6 34,700
Virginia 5.4 179,600 7.3 254,500 6.4 434,100
Washington 4.7 143,400 5.6 172,300 5.1 315,700
West Virginia 11.8 82,300 14.4 103,900 13.1 186,200
Wisconsin 5.5 126,400 5.7 132,800 5.6 259,200
Wyoming 6.2 14,100 6.9 15,100 6.5 29,200
United States 5.5 6,800,400 6.8 8,875,100 6.2 15,587,000
"
```

This formats our data so R can effectively read and format the states data
First we define define our pattern and format the COPD data
Second we create our data frame

```{r}
lines <- str_split(data_text, "\n")[[1]]
lines <- trimws(lines)
lines <- lines[lines != ""]
lines <- lines[-1]

# 4. Define the pattern
pattern <- "^([A-Za-z .()0-9]+?)\\s+(\\d+\\.\\d)\\s+([\\d,]+)\\s+(\\d+\\.\\d)\\s+([\\d,]+)\\s+(\\d+\\.\\d)\\s+([\\d,]+)$"
rows <- str_match(lines, pattern)
rows <- rows[!is.na(rows[,1]), ]

# 7. Create the data frame
df <- as.data.frame(rows[,2:8])
colnames(df) <- c("State","Percent_Male","Count_Male","Percent_Female","Count_Female","Percent_Total","Count_Total")
df <- df %>%
  # Filter out rows i don't want to plot
  filter(State != "United States") %>%
  filter(State != "District of Columbia") %>% # <-- NEW FIX: Remove D.C.
  # Convert columns to the right type
  mutate(across(starts_with("Percent"), as.numeric),
         across(starts_with("Count"), ~ as.numeric(str_replace_all(.x, ",", ""))),
         
         State = str_replace(State, "\\s*\\(.*\\)", ""), # Removes (2020)
         State = str_trim(State))                       # Removes whitespace
```

Creation of the COPD gender maps and colors
We first join the map data and create the color scale
We then map the data

```{r}
# 1. map data
states_map <- map_data("state")

# 2. heat map data to join
heatmap_data <- df %>%
  select(State, Percent_Female, Percent_Male) %>%
  pivot_longer(cols = starts_with("Percent_"),
               names_to = "Group",
               values_to = "Percent") %>%
  mutate(Group = str_replace(Group, "Percent_", ""),
         region = tolower(State)) 
map_plot_data <- left_join(states_map, heatmap_data, by = "region", relationship = "many-to-many")

map_plot_data %>%
  filter(!is.na(Group)) %>% 
  ggplot(aes(x = long, y = lat, group = group, fill = Percent)) +
  
  geom_polygon(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Percent") +
  facet_wrap(~ Group) + 
  labs(title = "COPD Prevalence by U.S. State and Gender") +
  theme_void() + 
  coord_map() + 
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    strip.text = element_text(size = 14, face = "bold")
  
  )
```

Figure 1. COPD prevalence is compared between males and females in the continental United States. Females show a slightly higher average of COPD.

This code creates the difference between sexes map.
We first differentiate male and female.
Second we plot the difference.
Lastly the color scale is created and the map is printed.

```{r}

states_map <- map_data("state") 
difference_data <- df %>%
  mutate(
    Percent_Difference = Percent_Female - Percent_Male,
    region = tolower(State) # <-- This creates the join key
  ) %>%
  select(region, Percent_Difference)
map_diff_data <- left_join(states_map, difference_data, by = "region")
ggplot(map_diff_data, aes(x = long, y = lat, group = group, fill = Percent_Difference)) +
  geom_polygon(color = "white", linewidth = 0.1) + # Draw the states
  # 5. Use the DIVERGING color scale
  scale_fill_gradient2(
    name = "Percent Difference\n(Female - Male)",
    low = "blue",      # States where males are higher (negative)
    mid = "white",     # States where rates are equal (zero)
    high = "red",      # States where females are higher (positive)
    midpoint = 0       # We center the scale at zero
  ) +
  
  labs(title = "Difference in COPD Prevalence (Female vs. Male)") +
  theme_void() + # Use a clean theme
  coord_map() +  # Use correct map projection
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

Figure 2. COPD percent difference for females compared to males in shown, with most states having a higher percentage of COPD prevalence in females.


# AIR QUALITY INDEX
```{r include = FALSE}
# more packages
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(viridis)
library(maps)
library(readr) # Added to read the CSV
```


```{r include=FALSE}
aq_df <- read_csv("air-quality-by-state-2025.csv")

print("--- Air Quality Data ---")

```

This code chunk creates the overall AQI map.
We begin by reformatting the raw data.
Second we join the reformatted data to the mapping package.
Lastly we print the map

```{r}
states_map <- map_data("state")

aq_data_to_plot <- aq_df %>%
  # Use the correct column names: `state` and `AirQuality_AirQualityIndexViaUSA_num_YearFree`
  # We also rename the long AQI column to `Overall_AQI` to make it easier to use
  select(State = state, Overall_AQI = `AirQuality_AirQualityIndexViaUSA_num_YearFree`) %>% 
  mutate(region = tolower(State)) # <-- This creates the join key

map_plot_data <- left_join(states_map, aq_data_to_plot, by = "region")


map_plot_data <- map_plot_data %>%
  filter(!is.na(Overall_AQI))

ggplot(map_plot_data, aes(x = long, y = lat, group = group, fill = Overall_AQI)) +
  geom_polygon(color = "white", linewidth = 0.1) + # Draw the states
  
  scale_fill_viridis_c(name = "Overall AQI") +
  
  labs(title = "Overall Air Quality Index (AQI) by U.S. State (2025)") +
  theme_void() + # Use a clean theme
  coord_map() +  # Use correct map projection
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

Figure 3. Overall AQI is shown for each of the continental United States, averaged over the year 2025. All states land generally within an acceptable range for daily intake, with some on the higher side indicating a lower AQI.
``` {r include = FALSE, eval = FALSE}
> ggplot(df, aes(x = sqrt(Percent_Total))) +
+     geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") +
+     labs(
+         title = "Normalized Distribution of COPD Prevalence",
+         x = "Square Root of Overall COPD Rate",
+         y = "Number of States"
+     ) +
+     theme_minimal() +
+     theme(
+         plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
+     )
> # We must use backticks for the column name
> aqi_column_name <- "AirQuality_AirQualityIndexViaUSA_num_YearFree"
> 
> # Use mutate() to create the new transformed column
> aq_df_normalized <- aq_df %>%
+     mutate(
+         # 1. Find the max value (k)
+         k = max(!!sym(aqi_column_name)) + 1,
+         
+         # 2. "Invert" (Reflect) and "SQRT"
+         # The formula is sqrt(k - x)
+         AQI_Normalized = sqrt(k - !!sym(aqi_column_name))
+     )
> 
> # Now, plot the new normalized column
> ggplot(aq_df_normalized, aes(x = AQI_Normalized)) +
+     geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black") + 
+     labs(
+         title = "Normalized Distribution of AQI (Reflected + SQRT)",
+         x = "Square Root of (Max AQI - AQI)",
+         y = "Number of States"
+     ) +
+     theme_minimal() +
+     theme(
+         plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
+     )


## combine the data
```

This code normalizes and reformats our data. 
We begin by loading needed libraries.
Then we load in our data and format to recombine them.
Lastly we normalize our data using the SQRT method.

```{r}
# 1. Make sure you have the 'readr' library
library(readr)
library(dplyr)
library(tidyr) # For pivoting later
library(ggplot2) # For the scatter plot

# 2. Reload two data frames 
aq_df <- read_csv("air-quality-by-state-2025.csv")

# 3. Join the two data frames by state
# We need to make sure the state names match.
# 'df' has "State" (e.g., "Alabama")
# 'aq_df' has "state" (e.g., "Alabama")

combined_df <- inner_join(df, aq_df, by = c("State" = "state"))


aqi_column_name <- "AirQuality_AirQualityIndexViaUSA_num_YearFree"

combined_df <- combined_df %>%
  mutate(
    # A. Normalize COPD rates (Square Root)
    Percent_Total_Norm = sqrt(Percent_Total),
    Percent_Male_Norm = sqrt(Percent_Male),
    Percent_Female_Norm = sqrt(Percent_Female),
    
    # B. Normalize AQI (Reflect + Square Root)
    k = max(!!sym(aqi_column_name)) + 1,
    AQI_Normalized = sqrt(k - !!sym(aqi_column_name))
  )


```

This code plots our SQRT and SQRT reflected data.
We first create the histogram and colors.
Second we create the axes

```{r}
ggplot(df, aes(x = Percent_Total)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Overall COPD Prevalence by State",
    x = "Overall COPD Rate (%)",
    y = "Number of States"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

Figure 4. Overall COPD rate percent is shown to be around 6, with some variation among other states.

```{r}
ggplot(aq_df, aes(x = `AirQuality_AirQualityIndexViaUSA_num_YearFree`)) +
  # AQI values are clustered, so a smaller binwidth is better
  geom_histogram(binwidth = 2, fill = "lightgreen", color = "black") +
  labs(
    title = "Distribution of Overall AQI by State (2025)",
    # We can set a cleaner label for the x-axis
    x = "Overall Air Quality Index (AQI)",
    y = "Number of States"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

```

Figure 5. Overall AQI is shown to be around 44, with a wide variation among other states.

## PEARSON TEST & GLM
```{r include = FALSE, eval = FALSE}
# This code block is for testing data normalization.
# It is currently not being run (eval = FALSE).

# 1. Plot histogram for normalized COPD data
ggplot(df, aes(x = sqrt(Percent_Total))) +
     geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") +
     labs(
         title = "Normalized Distribution of COPD Prevalence",
         x = "Square Root of Overall COPD Rate",
         y = "Number of States"
     ) +
     theme_minimal() +
     theme(
         plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
     )

# 2. Normalize the AQI data
# We must use backticks for the column name
aqi_column_name <- "AirQuality_AirQualityIndexViaUSA_num_YearFree"

# Use mutate() to create the new transformed column
aq_df_normalized <- aq_df %>%
     mutate(
         # 1. Find the max value (k)
         k = max(!!sym(aqi_column_name)) + 1,
         
         # 2. "Invert" (Reflect) and "SQRT"
         # The formula is sqrt(k - x)
         AQI_Normalized = sqrt(k - !!sym(aqi_column_name))
     )

# 3. Plot the new normalized AQI column
ggplot(aq_df_normalized, aes(x = AQI_Normalized)) +
     geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black") + 
     labs(
         title = "Normalized Distribution of AQI (Reflected + SQRT)",
         x = "Square Root of (Max AQI - AQI)",
         y = "Number of States"
     ) +
     theme_minimal() +
     theme(
         plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
     )

```

This code runs our combined data through a pearson correlation test, and a GLM model validate our data significance

```{r}
# Run a Pearson's Correlation Test
total_corr_test <- cor.test(combined_df$AQI_Normalized, combined_df$Percent_Total_Norm)

print(total_corr_test)

# Run the Simple Linear Regression (lm)
total_lm_model <- lm(Percent_Total_Norm ~ AQI_Normalized, data = combined_df)

print(summary(total_lm_model))
```

This code runs the same code as above between sexes

```{r}
# Correlation for MALES
male_corr_test <- cor.test(combined_df$AQI_Normalized, combined_df$Percent_Male_Norm)
print(male_corr_test)

# Correlation for FEMALES
female_corr_test <- cor.test(combined_df$AQI_Normalized, combined_df$Percent_Female_Norm)
print(female_corr_test)

```

This code runs a GLM model against each sex to determine if there is statistical significance between sexes

```{r}
long_df <- combined_df %>%
  select(State, AQI_Normalized, Percent_Male_Norm, Percent_Female_Norm) %>%
  pivot_longer(
    cols = c("Percent_Male_Norm", "Percent_Female_Norm"),
    names_to = "Gender",
    values_to = "COPD_Norm"
  ) %>%
  mutate(Gender = str_replace(Gender, "Percent_", "")) %>%
  mutate(Gender = str_replace(Gender, "_Norm", ""))
gender_interaction_model <- lm(COPD_Norm ~ AQI_Normalized * Gender, data = long_df)

summary(gender_interaction_model)
```


This code generates the final trendline plot and plots the data from the GLM.

```{r final plot}
#This code chunk was troubleshooted and enhanced with AI
ggplot(combined_df, aes(x = AQI_Normalized, y = Percent_Total_Norm)) +
  geom_point(color = "black", alpha = 0.6, size = 2.5) +
  
  # Add the regression line
  geom_smooth(method = "lm", color = "darkblue", fill = "lightblue") +
  
  labs(
    title = "Overall Relationship: Air Quality vs. COPD Rates",
    x = "Normalized AQI Score \n(← Worse Air Quality | Better Air Quality →)",
    y = "Normalized Total COPD Rate"
  ) +
  
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text = element_text(color = "black")
  )

```

Fig. 5 The graph visualizes the results from the Pearson correlation and the GLM. In this plot it is evident that there is a correlation between AQI and COPD rates



# DISCUSSION
  Our primary analysis revealed a strong statistical correlation between the Air Quality Index (AQI) and rates of Chronic Obstructive Pulmonary Disease (COPD) (p=0.0179). This leads us to reject the null hypothesis and support our initial hypothesis, confirming that states with poorer average air quality have higher rates of COPD. Although the relationship is significant, the effect size is moderate. Our model produced an R² value of 0.1112, indicating that AQI accounts for approximately 11.1% of the variation in COPD rates. This implies that other unaccounted confounding factors, such as smoking, occupation, and genetics, likely play a more substantial role.

We also examined the impact of AQI on biological sexes. The Pearson correlation test indicated a slightly stronger relationship for females (p = 0.0099) compared to males (p = 0.0474); however, our generalized linear model revealed that this difference was not statistically significant. The interaction term between gender and air quality yielded a p-value of 0.512, suggesting that the regression slopes for males and females are statistically indistinguishable. This indicates that poor air quality is an equal-opportunity risk factor; as air quality declines, COPD rates increase at similar rates for both sexes.


Limitations:

  Firstly, relying on aggregate-level state data may not accurately represent the average air quality experienced by individuals. For instance, in Utah, much of the poor air quality is concentrated in the two major valley centers, while residents outside these valleys often enjoy significantly better air quality. Secondly, migration likely has a substantial impact on the data. An individual might develop COPD in a state with poor air quality and then relocate to a state with better conditions, which could distort the overall findings.

# CONCLUSION

  Our data indicates a correlation between areas with higher Air Quality Index (AQI) numbers and increased rates of Chronic Obstructive Pulmonary Disease (COPD). With this understanding, we can better inform residents in regions with elevated AQI levels to take necessary precautions to mitigate the risk of developing COPD. Recommended practices include minimizing outdoor activities during times of poor air quality, utilizing effective air filters in their homes, and curtailing the use of machinery that contributes to elevated AQI levels. Leveraging this knowledge could prove invaluable in reducing COPD rates across the United States and helping individuals regain the quality of life they deserve.

# REFERENCES
American Lung Association. (2023). COPD prevalence rates and counts by state and gender. Retrieved November 18, 2025, from https://www.lung.org/research/trends-in-lung-disease/copd-trends-brief/data-tables/copd-prevalence-rates-by-state-gender

Duan, R.R., Hao, K., & Yang, T. (2020). Air pollution and chronic obstructive pulmonary disease. Chronic Diseases and Translational Medicine, 6(4), 260–269. https://doi.org/10.1016/j.cdtm.2020.05.004

Google. (2025). Gemini (Version Jan 2025) [Large language model]. Accessed `r Sys.Date()`.

OpenAI. (2025). ChatGPT (Jan 2025 version) [Large language model]. Accessed `r Sys.Date()`.

World Population Review. (2025). Air Quality by State 2025. https://worldpopulationreview.com/state-rankings/air-quality-by-state
